---
title: "Worksheet 01 - Intro to R programming - NCBS MSc WL (Answers)"
author: 'Akshay Surendra, Anand M Osuri'
date: '23 - September - 2020'
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE,eval = TRUE, echo = TRUE)
```

Convert the following tasks to R commands and run them:

  1. Create a vector `g` that stores even numbers from 10 to 20 in a sequence
```{r}
g <- seq(from = 10,to = 20,by = 2)
```
  
  2. Find the logarithm (base-2) of each element of vector `g` in just one line of code (Clue: *vectorize*)
```{r}
logG <- log2(x = g) # built-in function specifically for log base-2
logG
logG <-log(x = g,base = 2) # generic log function where you can change the base
```
  
  
  3. Can you implement 1) and 2) above in one line of code?
  (Clue: nested functions)
```{r}
logG <- log2(x = seq(from = 10,to = 20,by = 2))
```
  
    
  4. Run the following code *snippet* (another word for a block of code) in your R script \-
```{r}
  matrix(data = rnorm(n = 32,mean = 0,sd = 2),nrow = 4,ncol = 8)
```
  
\underline{Next class}: read more about the `rnorm()` function and what it does - this is not required to answer the questions below
  
  5. Save the above matrix in an object and use appropriate indices to extract and print the following:
```{r}
obj1 <- matrix(data = rnorm(n = 32,mean = 0,sd = 2),nrow = 4,ncol = 8)#run once
obj1
```
  
    a. even-numbered columns
```{r}
# a
obj1[,c(2,4,6,8)]
```
    
    b. odd-numbered rows
```{r}
# b
obj1[c(1,3),]
```
    
    c. both even-numbered columns AND odd-numbered rows
```{r}
# c
obj1[c(1,3),c(2,4,6,8)]
```

  6. In the above matrix, *check* if the following statements are true or false (Clue\- use one or more of these three operators: relational/logical/assignment):
  
  a. Is the 1st element of row 1 less than the 8th element of row 1?
```{r}
obj1[1,1] < obj1[1,8]
```

  b. Is the 4th element of column 1 more than the 6th element of column 1?
  (Report any result you get, error or otherwise)
```{r}
# obj1[4,1] > obj1[6,1] ## gives you an error - subscript out of bounds
```
    
  7. Learn what the function `substr()` does using the help command (and also the internet) and answer the following questions:
  
  a. Create a character vector object with 6 elements, and each element must hold the name of a bird native to your country (data type: ?)
```{r}
obj2 <- c("HoodedCrow","LaughingDove","Redwing","HazelGrouse","Gadwall","AmurFalcon")
obj2
```
  
  b. Identify and print the first 2 letters of each element in this vector
```{r}
first2_1 <- substr(x = obj2[1],start = 1,stop = 2) # includes the starting and ending index
first2_2 <- substr(x = obj2[2],start = 1,stop = 2)
first2_3 <- substr(x = obj2[3],start = 1,stop = 2)
first2_4 <- substr(x = obj2[4],start = 1,stop = 2)
first2_5 <- substr(x = obj2[5],start = 1,stop = 2)
first2_6 <- substr(x = obj2[6],start = 1,stop = 2)

first2_1
first2_2
first2_3
first2_4
first2_5
first2_6

```
  
  c. If you used 6 lines of code to implement 7-b, can you do it in 1 line? (clue \- *vectorize*)
```{r}
first2 <- c(NA,NA,NA,NA,NA,NA) # not required but good practice to define length first
first2 <- NULL # empty vector, akin to saying, d <- data.frame()
first2 <- substr(x = obj2,start = 1,stop = 2)
first2

```
